Laurens de Graaf

### 2015-05-04 writeup/paper.tex
Two sigplan pages are done, but the citations are a mess

\lstnewenvironment{code}[1][]%
  {\noindent\medskip 
     \lstset{basicstyle=\sffamily\footnotesize,frame=single,#1}}
  {}

@@

In this section, we give a brief sketch of the implementation approach we take
to construct the STG Machine in RPython, discussing the general structure of the
implementation. To begin, we sketch the directory structure and describe each file
there:

\begin{itemize}
\item \lstinline{laurens/}
  \begin{itemize}
    \item \lstinline{config.py} -- Configuration file (currently unused)
    \item \lstinline{debug.py}  -- Debug utilities, primarily for printing
                                   machine states during execution
    \item \lstinline{op.py}     -- The operation nodes, which each contain
                                   a \lstinline{step} operation that performs
                                   a machine reduction.
    \item \lstinline{parse.py}  -- The sketch of an input parser (currently 
                                   unstarted)
    \item \lstinline{stg.py}    -- The main loop of the implementation
    \item \lstinline{tests.py}  -- A set of tests written in terms of AST
                                   nodes for the machine; this serves as
                                   an alternative entry point
  \item \lstinline{laurens/ast/}
    \begin{itemize}
      \item \lstinline{ast.py}  -- Contains the AST nodes that make up expressions;
                                   part of \lstinline{op.py} should likely be
                                   refactored into this file
      \item \lstinline{cont.py} -- Continuations that are pushed to the Return Stack 
    \end{itemize}
  \item \lstinline{laurens/data/}
    \begin{itemize}
      \item \lstinline{argstack.py} -- Stack for arguments to functions
      \item \lstinline{closure.py}  -- Data structure to represent closures
      \item \lstinline{config.py}   -- Configuration file (currently unused)
      \item \lstinline{heap.py}     -- Heap structure
      \item \lstinline{heapobj.py}  -- Objects that live on the heap
      \item \lstinline{retstack.py} --
      \item \lstinline{updframe.py} --
      \item \lstinline{updstack.py} -- 
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\subsection{The Configuration Structure}
We begin by briefly sketching the machine configuration, which contains
six individual pieces:
\[\langle op; as; rs; us; h; \sigma\rangle\]
Unsurprisingly, each component plays a unique role in the machine:
\begin{itemize}
  \item $op$ contains the operation for the reduction to perform;
  \item $as$ contains the argument stack, which maintains arguments
        during applications;
  \item $rs$ contains the return stack, which manages return points
        (primarily ``case'' expressions);
  \item $us$ contain the update stack, used for handling thunk updates;
  \item $h$ contains the heap, where memory addresses map to closures;
  \item $\sigma$ contains the \emph{global environment} of bindings.
\end{itemize}
While \citet{spj:stgmachine} observe that many of these stacks may
be combined during compilation to C as an optimization, we maintain
this configuration explicitly in our implementation:
\begin{code}
class Config(object):
  def __init__(self, code, args, ret,  
                     upd, heap, genv):
    self.code       = code
    self.arg_stack  = args
    self.ret_stack  = ret
    self.upd_stack  = upd
    self.heap       = heap
    self.global_env = genv
\end{code}
We take this approach for two reasons. First, we are working from the
assumption that RPython should be able to properly handle the interpretation
structure of the STG machine, and thus these individual stacks should not
cause critical, performance-inhibiting overhead.
Second, RPython performs immense amounts of type inference and combining
these stacks would require that we carefully package each different node
in a new object to indicate which of the original stacks it lived on,
thus adding another level of indirection that the compiler must 
overcome\footnote{There are other mechanisms to avoid this discussed
in the original work, but they require subtle bit manipulation that
would need to be faked in a higher-level implementation.}.

This automatic type distinction has a further problem, however: we need
identical classes for each of the \lstinline{arg_stack}, \lstinline{ret_stack},
and \lstinline{upd_stack} so that RPython may infer each to a different,
appropriate type. Indeed, we would suggest that this is one of the largest
hurdles in constructing a project such as this in RPython: there is a non-zero
amount of code duplication required to simply exist in the RPython ecosystem.

@@

>>>>

\lstnewenvironment{code}[1][]%
  {\noindent\medskip 
     %\lstset{basicstyle=\sffamily\footnotesize,frame=single,#1}}
     \lstset{basicstyle=\sffamily\footnotesize,#1}}
  {}

@@

In this section, we give a brief sketch of the implementation approach we take
to construct the STG Machine in RPython, discussing the general structure of the
implementation. To begin, we sketch the directory structure and describe each file
in Figure~\ref{fig:dir} to outline the overall project.

\begin{figure}
\begin{itemize}
\item \lstinline{laurens/} -- Top-level directory
  \begin{itemize}
    \item \lstinline{config.py} -- Configuration file (currently unused)
    \item \lstinline{debug.py}  -- Debug utilities, primarily for printing
                                   machine states during execution
    \item \lstinline{op.py}     -- The operation nodes, which each contain
                                   a \lstinline{step} operation that performs
                                   a machine reduction.
    \item \lstinline{parse.py}  -- Input parser (currently unstarted)
    \item \lstinline{stg.py}    -- The main loop of the implementation
    \item \lstinline{tests.py}  -- A set of tests written in terms of AST
                                   nodes for the machine; this serves as
                                   an alternative entry point
  \item \lstinline{laurens/ast/}
    \begin{itemize}
      \item \lstinline{ast.py}  -- AST nodes that make up expressions;
                                   part of \lstinline{op.py} should likely be
                                   refactored into this file
      \item \lstinline{cont.py} -- Continuations for the  Return Stack 
    \end{itemize}
  \item \lstinline{laurens/data/}
    \begin{itemize}
      \item \lstinline{argstack.py} -- Stack for arguments to functions
      \item \lstinline{closure.py}  -- Data structure to represent closures
      \item \lstinline{config.py}   -- Machine Configuration
      \item \lstinline{heap.py}     -- Heap structure
      \item \lstinline{retstack.py} -- Stack for return continuations
      \item \lstinline{updframe.py} -- Update frame tuple
      \item \lstinline{updstack.py} -- Stack for update frames
    \end{itemize}
  \end{itemize}
\end{itemize}
\caption{\label{fig:dir}The directory structure of Laurens.}
\end{figure}

\subsection{The Machine Configuration}

Before we discuss the implementation details, we describe the
STG Machine. This abstract machine uses a six-tuple configuration:
\[\langle op; as; rs; us; h; \sigma\rangle\]
Unsurprisingly, each component plays a unique role in the machine:
\begin{itemize}
  \item $op$ contains the operation for the reduction to perform;
  \item $as$ contains the argument stack, which maintains arguments
        during applications;
  \item $rs$ contains the return stack, which manages return points
        (primarily ``case'' expressions);
  \item $us$ contain the update stack, used for handling thunk updates;
  \item $h$ contains the heap, where memory addresses map to closures;
  \item $\sigma$ contains the \emph{global environment} of bindings.
\end{itemize}
While \citet{spj:stgmachine} observe that many of these stacks may
be combined during compilation to C as an optimization, we maintain
this configuration explicitly in our implementation:
\begin{code}
class Config(object):
  def __init__(self, code, args, ret,  
                     upd, heap, genv):
    self.code       = code
    self.arg_stack  = args
    self.ret_stack  = ret
    self.upd_stack  = upd
    self.heap       = heap
    self.global_env = genv
\end{code}
We take this approach for two reasons. First, we are working from the
assumption that RPython should be able to properly handle the interpretation
structure of the STG machine, and thus these individual stacks should not
cause critical, performance-inhibiting overhead.
Second, RPython performs immense amounts of type inference and so we use
separate stack implementations, as we discuss further in
Section~\ref{sec:typeinf}.

\subsection{The Main Loop}

\begin{figure}
\begin{code}
def terminateHuh(config):
  return 
    (((config.code.op == op.ReturnConOp) or 
      (config.code.op == op.ReturnIntOp)) 
     and config.ret_stack.empty()
     and config.upd_stack.empty())

def loop(config):
  while not terminateHuh(config):
    jitdriver.jit_merge_point(code = 
                                config.code)
    config = config.code.step(config)
  return config
\end{code}
\caption{\label{fig:loop}The main loop and termination test.}
\end{figure}

The \lstinline{stg.py} file serves as the main loop, driven by operation steps
in \lstinline{op.py}. Everything else supports this main interaction.  
Our overall computational mechanism is presented in Figure~\ref{fig:loop}. The
main file contains a loop driver that steps the configuration until there is no
more to do. At each step, we call \lstinline{step} on the current configuration,
receiving a new one. This computation proceeds until we terminate with either
an integer or a constructor and both the update and return stacks are empty.

The operations are broken up by the STG machine into four possible operations:
evaluation, entering, returning a construct, and returning an integer, given
as \emph{Eval $e$ $\rho$}, \emph{Enter $a$}, \emph{ReturnCon $c$ $xs$}, and
\emph{ReturnInt $i$} in the reduction semantics. We implement each of these
as an \emph{operation node} in \lstinline{op.py}, wherein each has a
\lstinline{step} method that expects a configuration using that operator
and dispatches appropriately, returning the new machine configuration.

@@

\subsection{Unique Stack Implementations}\label{sec:typeinf}

As previously state, RPython attempts exhaustive type inference for structures
during compilation. This approach has forced us into using three different
stack implementations, all identical in code; we would like to use a single
stack defined as \lstinline{data/stack.py}, but these authors could not find
a way to circumvent the type inference to allow such reuse.
Indeed, we would suggest that this is one of the largest
hurdles in constructing a project such as this in RPython: there is a non-zero
amount of code duplication required to simply exist in the RPython ecosystem.

Furthermore, the later sections of \citet{spj:stgmachine} describe an
optimization for combining stacks that would suggest that we can combine these
stacks by carefully packaging each different node in a new object to indicate
which of the original stacks it lived on\footnote{There are other mechanisms to
avoid this discussed in the original work, but they require subtle bit
manipulation that would need to be faked in a higher-level implementation.}. 
Unfortunately, this would add one further level of indirection for the compiler
and further obfuscate our implementation, and it is unclear if we could gain
speed for the effort.  Further experimentation should be done, here, however.

@@

Finally, we discuss our timing results, and must report a negative one:
the final implementation is \emph{slower} when we perform tracing on
a sample program computing the 30th Fibonacci number. This is almost certainly
the result of our na\"{i}ve tracing mechanism, wherien we consider every node as
a possible merge point for the JIT, introducing immense tracing overhead to the
computation. Thus it is unclear if Tracing is suitable for the STG machine
because we have not performed effort to fine-tune our implementation.

Beyond this result, we still have much to do:
\begin{itemize}
\item Write a full parser that will construct AST nodes from reasonable input;
\item Write a more extensive test suite to examine and use constructors,
      \lstinline{let}, \lstinline{letrec}, laziness, etc.; this will be much
      easier once a parser is written;
\item Support a full range of built-in functions, including I/O and additional
      numeric operators;
\item Replace the dictionary-based representation of environments with a
      more efficient data structure
\item Fine-tune the JIT heuristics to gain speed.
\end{itemize}
